---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.path='figures\\')
```

## Loading and preprocessing the data
You will need to have extracted the document activity.csv into a folder named activity.

```{r extract}
activity_data <- read.csv("./activity/activity.csv")
head(activity_data)
summary(activity_data)
```


## What is mean total number of steps taken per day?

#### Daily Steps Distribution
```{r hist steps, fig.width=10 }
require(ggplot2)
ggplot(data = activity_data) + 
  geom_histogram(aes(y = steps, x = date), stat = "identity") +
  labs(x = "Date", y = "Average Number of Steps") + 
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 0.8, vjust = 0.4)) +
  ggtitle("Histogram of Total Number of Steps Taken per Day")

stepsByDate <- aggregate(activity_data$steps, by = list(date = activity_data$date), FUN = sum )
```
#### Average Daily Steps
```{r}
mean(stepsByDate$x, na.rm = TRUE)
```

#### Median Daily Steps
```{r}
median(stepsByDate$x, na.rm = TRUE)
```


## What is the average daily activity pattern?

#### Time Series Plot
```{r fig.width=10}
stepsByInterval <- aggregate(activity_data$steps, by = list(interval = activity_data$interval), FUN = mean, na.rm = TRUE)
ggplot(stepsByInterval) +
  geom_line(aes(y = x, x = interval)) +
  labs(y = "Average number of Steps") +
  ggtitle("Average Number of Steps Taken in Each Time Interval of the Day")
```

#### Peak Activity Period

```{r fig.width=10}
colnames(stepsByInterval)[2] <- "steps"
##this is the interval with the highest average number of steps
stepsByInterval[which.max(stepsByInterval$steps),]
```
## Imputing missing values

#### Number of rows with Missing (NA) Values
```{r}
length(which(complete.cases(activity_data) == FALSE))
```

#### Filling in Blanks
```{r}
##Filling in missing "NA" values with the mean value for the interval.
newsteps <- merge(x = activity_data, y = stepsByInterval, by = "interval", all.x = TRUE)
newsteps$steps.x[is.na(newsteps$steps.x)] <- newsteps$steps.y[is.na(newsteps$steps.x)]
newsteps$steps.x <- as.integer(newsteps$steps.x)
newsteps <- newsteps[, 1:3]
colnames(newsteps)[2] <- "steps"
newsteps <- newsteps[, c(2,3,1)]
```

#### Daily Steps Distribution - NA's replaced with mean
```{r hist steps na, fig.width=10}
require(ggplot2)
ggplot(data = newsteps) + 
  geom_histogram(aes(y = steps, x = date), stat = "identity") +
  labs(x = "Date", y = "Average Number of Steps") + 
  theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 0.8, vjust = 0.4)) +
  ggtitle("Histogram of Total Number of Steps Taken per Day")

stepsByDate <- aggregate(newsteps$steps, by = list(date = newsteps$date), FUN = sum )
```
#### Average Daily Steps
```{r}
mean(stepsByDate$x, na.rm = TRUE)
```

#### Median Daily Steps
```{r}
median(stepsByDate$x, na.rm = TRUE)
```
#### Results of Imputing Data
While it gives our graph more continuity, it doesn't greatly affect the overall picture.

## Are there differences in activity patterns between weekdays and weekends?
#### Add a column indicating week or weekend
```{r}
newsteps$dayofweektype <- as.factor(
  ifelse(
    weekdays(as.POSIXlt( newsteps$date)) %in% c("Saturday", "Sunday"), 
    "Weekend", 
    "Weekday"
    )
  )
```

#### Steps per Interval - Weekdays vs Weekends 
```{r}
newstepsByInterval <- aggregate(newsteps$steps, by = list(interval = newsteps$interval, dayofweektype = newsteps$dayofweektype), FUN = mean, na.rm = TRUE)
ggplot(newstepsByInterval) +
  geom_line(aes(y = x, x = interval)) +
  labs(y = "Average number of Steps") +
  ggtitle("Average Number of Steps Taken in Each Time Interval of the Day") +
  facet_wrap(~dayofweektype, ncol = 1)
```

